### 2020-10-23 Frequency-of-Travel-by-Mode-COVID-19-Bangladesh/Frequency-of-Travel-by-Mode-COVID-19-Bangladesh.Rmd
v2 Corrected missing reference after proof-reading by Shaila, and rendered README.Rmd

Like other surveys conducted during COVID-19 [e.g., @Astroza2020mobility], this is a convenience sample. Convenience samples have the advantage of being quick and cost-efficient and, despite their lower generalizability, when they are homogeneous they are useful to address rapidly emerging questions [Fricker2002advantages; @Jager2017convenience]. This particular data set has an over-representation of the 16-30 years age group, whereas 37% of the country's population belongs to this group [@Bangladesh2015]. Also, 66% of the respondents are male and 32% are females compared to country's 49:51 male-female ratio [@Bangladesh2015]. These characteristics of the sample are expected in terms of online surveys as young adults are more familiar with online platforms, especially in the context of a developing country context. Also, in the case of Bangladesh, collecting a representative sample remains difficult because of overall public reluctance to participate, with women's strong unwillingness to share their information to unknown parties in particular in face-to-face interviews [see @Jamal2020perceptions]. In fact, compared to  Jamal et al. [-@Jamal2020perceptions; Table 1], who had only 15% of responses from females, the online survey used in the present research achieved greater penetration among women.

>>>>

Like other surveys conducted during COVID-19 [e.g., @Astroza2020mobility], this is a convenience sample. Convenience samples have the advantage of being quick and cost-efficient and, despite their lower generalizability, when they are homogeneous they are useful to address rapidly emerging questions [@Fricker2002advantages; @Jager2017convenience]. This particular data set has an over-representation of the 16-30 years age group, whereas 37% of the country's population belongs to this group [@Bangladesh2015]. Also, 66% of the respondents are male and 32% are females compared to country's 49:51 male-female ratio [@Bangladesh2015]. These characteristics of the sample are expected in terms of online surveys as young adults are more familiar with online platforms, especially in the context of a developing country context. Also, in the case of Bangladesh, collecting a representative sample remains difficult because of overall public reluctance to participate, with women's strong unwillingness to share their information to unknown parties in particular in face-to-face interviews [see @Jamal2020perceptions]. In fact, compared to  Jamal et al. [-@Jamal2020perceptions; Table 1], who had only 15% of responses from females, the online survey used in the present research achieved greater penetration among women.

### 2020-10-23 Frequency-of-Travel-by-Mode-COVID-19-Bangladesh/Frequency-of-Travel-by-Mode-COVID-19-Bangladesh.tex
v2 Corrected missing reference after proof-reading by Shaila, and rendered README.Rmd

Like other surveys conducted during COVID-19 (e.g., Astroza et al.,
2020), this is a convenience sample. Convenience samples have the
advantage of being quick and cost-efficient and, despite their lower
generalizability, when they are homogeneous they are useful to address
rapidly emerging questions (Fricker2002advantages; Jager et al., 2017).
This particular data set has an over-representation of the 16-30 years
age group, whereas 37\% of the country's population belongs to this
group (Bangladesh Bureau of Statistics, 2015). Also, 66\% of the
respondents are male and 32\% are females compared to country's 49:51
male-female ratio (Bangladesh Bureau of Statistics, 2015). These
characteristics of the sample are expected in terms of online surveys as
young adults are more familiar with online platforms, especially in the
context of a developing country context. Also, in the case of
Bangladesh, collecting a representative sample remains difficult because
of overall public reluctance to participate, with women's strong
unwillingness to share their information to unknown parties in
particular in face-to-face interviews (see Jamal et al., 2020). In fact,
compared to Jamal et al.~{[}(2020); Table 1{]}, who had only 15\% of
responses from females, the online survey used in the present research
achieved greater penetration among women.

@@

>>>>

Like other surveys conducted during COVID-19 (e.g., Astroza et al.,
2020), this is a convenience sample. Convenience samples have the
advantage of being quick and cost-efficient and, despite their lower
generalizability, when they are homogeneous they are useful to address
rapidly emerging questions (Fricker and Schonlau, 2002; Jager et al.,
2017). This particular data set has an over-representation of the 16-30
years age group, whereas 37\% of the country's population belongs to
this group (Bangladesh Bureau of Statistics, 2015). Also, 66\% of the
respondents are male and 32\% are females compared to country's 49:51
male-female ratio (Bangladesh Bureau of Statistics, 2015). These
characteristics of the sample are expected in terms of online surveys as
young adults are more familiar with online platforms, especially in the
context of a developing country context. Also, in the case of
Bangladesh, collecting a representative sample remains difficult because
of overall public reluctance to participate, with women's strong
unwillingness to share their information to unknown parties in
particular in face-to-face interviews (see Jamal et al., 2020). In fact,
compared to Jamal et al.~{[}(2020); Table 1{]}, who had only 15\% of
responses from females, the online survey used in the present research
achieved greater penetration among women.

@@

\leavevmode\hypertarget{ref-Fricker2002advantages}{}%
Fricker, R.D., Schonlau, M., 2002. Advantages and disadvantages of
internet research surveys: Evidence from the literature. Field methods
14, 347--367.

### 2020-10-23 README.Rmd
v2 Corrected missing reference after proof-reading by Shaila, and rendered README.Rmd

---
title: An exploratory analysis of changes in trip-making frequency by mode during the COVID-19 emergency in Bangladesh
author:
  - name: Shaila Jamal
    email: jamals16@mcmaster.ca
    affiliation: McMaster University
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
 The COVID-19 pandemic has had a profound impact on mobility in every country and region around the world. Recent helps to illuminate some of the dimensions of change - however, the evidence is still scant in developing countries. The objective of this paper is to present an exploratory analysis of the changes in the frequency of trip-making by mode during the COVID-19 emergency in Bangladesh. The results confirm an overall loss of mobility in the form of reduced trip-making frequency by all modes, but the changes are not uniform across modes. In particular, there was a greater loss of mobility for bus, rickshaw, and CNG auto-rickshaw.

journal: "Transport Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
always_allow_html: true

header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

@@

Data used for this paper come from an online survey, "Exploring the potential of travel mode change behavior in the post-lock-down and post-pandemic (COVID-19) period" conducted during July - August 2020 in Bangladesh. The survey was available both in Bengali and English and disseminated through various electronic means along with a widespread social media campaign and promotion in various social media sites and groups. In addition to socio-economic and demographic information of the respondents, the survey covered different aspects of travel such as travel behavior before and during COVID-19, knowledge related to COVID-19, and opinions and perceptions regarding travel behavior after lock-down and post-pandemic situation.

For the purpose of this paper, we use two questions that provide information about trip-making frequency by eight modes of transportation. The modes are **car**, **ridesharing** (e.g., Uber, Pathao), **rickshaw**, **CNG auto-rickshaw** (a rickshaw-like vehicle powered by compressed natural gas), **bus**, motorcycle/scooter (hereafter just **motorcycle**), **walking**, and **bicycle** (there was an additional catch-all category **other** which we ignore here). Participants in the survey used the following levels to report their frequency of traveling by each mode both before and during COVID-19: _Never_, _Rarely_, _Once a week_, _2-3 trips per week_, _4-5 trips per week_, _Almost daily_. There are $n=800$ responses in the data set.

@@

Figure \ref{fig:column-plot-cases} shows the number of responses (out of 800) in each trip-making frequency class by mode of transportation. The white bars and gray bars are for travel before and during the pandemic, respectively. Considering travel before the pandemic, travel by rickshaw, and bus were relatively common for many respondents (few respondents reported _never_ using these modes). The mode most commonly used on a quotidian basis was walk. In contrast, respondents reported less frequent travel by car, ridesharing services, CNG auto-rickshaw, motorcycle, and bicycle. During the pandemic, we see that while there were reductions in mobility by car, motorcycle, and bicycle (with more respondents reporting never traveling by these modes), the changes were relatively minor.

```{r column-plot-cases-before-after, echo=FALSE, fig.width= 5, fig.cap="\\label{fig:column-plot-cases}Number of responses by trip-making frequency class and mode, before and during COVID-19"}
freq_by_mode_long %>%
  filter(Mode != "other") %>%
  ggplot(aes(x = frequency, y = Cases, fill = Period)) +
  geom_col(color = "black", position = "dodge") + 
  scale_fill_manual(values = c("before" = "white", "during" = "gray")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  facet_wrap(. ~ Mode, nrow = 4)
```

```{r table-transitions, echo=FALSE}
rbind(car_xtab, 
      rideshare_xtab,
      rickshaw_xtab, 
      cng_xtab,
      bus_xtab, 
      motorcycle_xtab,
      walk_xtab,
      bicycle_xtab) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:cross-tabulation} Cross-tabulation of counts of travel frequency by mode, before and during COVID-19") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>%
  pack_rows("car", 1, 6) %>%
  pack_rows("rideshare", 7, 12) %>%
  pack_rows("rickshaw", 13, 18) %>%
  pack_rows("CNG auto-rickshaw", 19, 24) %>%
  pack_rows("bus", 25, 30) %>%
  pack_rows("motorcycle", 31, 36) %>%
  pack_rows("walk", 37, 42) %>%
  pack_rows("bicycle", 43, 48)
```

The frequency of trip-making by other modes changed more noticeably: the frequency of travel by ridesharing services, rickshaw, CNG auto-rickshaw, and bus collapsed, with vastly more respondents reporting never using these modes during the pandemic than before. The frequency of walking trips also decreased (fewer respondents report walking almost daily), but the reductions in mobility were not so heavily concentrated at the bottom of the scale.

@@

Despite across-the-board losses of mobility, there appears to have been some adaptation that varied by mode of transportation. To illustrate, `r bus_xtab[6, 1]` respondents, or `r paste(round(bus_xtab[6, 1] / sum(bus_xtab[6,]) * 100, 2), "%", sep="")` of those who traveled by bus almost daily before, reported never using it during the pandemic. In contrast, only `r sum(bus_xtab[1, 2:6])` respondents, or `r paste(round(sum(bus_xtab[1, 2:6]) / sum(bus_xtab[,]) * 100, 2), "%", sep="")` of those who never used buses before started doing so during the pandemic. By way of comparison, `r paste(round(bicycle_xtab[6, 1] / sum(bicycle_xtab[6,]) * 100, 2), "%", sep="")` of respondents who cycled almost daily before the pandemic stopped doing so - but `r paste(round(sum(bicycle_xtab[1, 2:6]) / sum(bicycle_xtab[,]) * 100, 2), "%", sep="")` who never cycled before started doing so during the pandemic.

@@

# BICYCLE
circle.plot(bicycle_ptab2, "bicycle", nlines = -3)
```

The findings reported here provide important changes in mobility in a developing country during COVID-19. In general there was a loss of mobility across-the-board, but modes that require interaction with strangers (rideshare, bus, rickshaw, and CNG auto-rickshaw) were more affected. Travel by car, motorcycle, and bicycle were somewhat less affected, and there is evidence of some adoption of walking during the pandemic.

>>>>

---
output: github_document
always_allow_html: true
bibliography: Frequency-of-Travel-by-Mode-COVID-19-Bangladesh/References.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Changes in trip-making frequency by mode during the COVID-19 emergency: evidence from an online survey in Bangladesh

<!-- badges: start -->
<!-- badges: end -->

Shaila Jamal (jamals16@mcmaster.ca)  
Antonio Paez (paezha@mcmaster.ca)  
School of Geography and Earth Science  
McMaster University  
1280 Main Street West, Hamilton, ON L8S 4K1  
Canada

Paper submitted for peer review to Transport Findings

## Abstract

The COVID-19 pandemic has had a profound impact on mobility in every country and region around the world. Recent studies help to illuminate some of the dimensions of change - however, the evidence is still scant in developing countries. The objective of this paper is to present an exploratory analysis of the changes in the frequency of trip-making by mode during the COVID-19 emergency in Bangladesh. The analysis is based on an online sample conducted during the pandemic, and the results confirm an overall loss of mobility, especially among younger people, in the form of reduced trip-making frequency by all modes. In addition, the results suggest that changes are not uniform across modes, and in particular the loss of mobility was more pronounced for bus, rickshaw, and CNG auto-rickshaw. In contrast, there was some adoption of walking during the pandemic.

## Keywords

- COVID-19
- Mobility
- Trip-making frequency
- Modes of transportation
- Transition probabilities
- Circular plots

@@

Data used for this paper come from an online survey, "Exploring the potential of travel mode change behavior in the post-lock-down and post-pandemic (COVID-19) period", conducted during July - August 2020 in Bangladesh ($n=800$). The survey was disseminated through various electronic means along with a widespread social media campaign and promotion in various social media sites and groups. In addition to the respondents' socio-demographic information, the survey covered different aspects of travel such as travel behavior before and during COVID-19, knowledge related to COVID-19, and opinions and perceptions regarding travel behavior after lock-down and post-pandemic situation.

Like other surveys conducted during COVID-19 [e.g., @Astroza2020mobility], this is a convenience sample. Convenience samples have the advantage of being quick and cost-efficient and, despite their lower generalizability, when they are homogeneous they are useful to address rapidly emerging questions [@Fricker2002advantages; @Jager2017convenience]. This particular data set has an over-representation of the 16-30 years age group, whereas 37% of the country's population belongs to this group [@Bangladesh2015]. Also, 66% of the respondents are male and 32% are females compared to country's 49:51 male-female ratio [@Bangladesh2015]. These characteristics of the sample are expected in terms of online surveys as young adults are more familiar with online platforms, especially in the context of a developing country context. Also, in the case of Bangladesh, collecting a representative sample remains difficult because of overall public reluctance to participate, with women's strong unwillingness to share their information to unknown parties in particular in face-to-face interviews [see @Jamal2020perceptions]. In fact, compared to  Jamal et al. [-@Jamal2020perceptions; Table 1], who had only 15% of responses from females, the online survey used in the present research achieved greater penetration among women.

In terms of geography, respondents were asked to identify the name of the district they reside. The resulting sample has 48% of respondents who live in Dhaka. Although, there are some differences in availability of transportation services between cities, especially between Dhaka and cities outside Dhaka, the mode use frequency indicates that only 15% never used buses and 32% never used ride-hailing services before COVID-19, which means these modes were available to most of the respondents and a vast majority has used them to some extent before the lock-down.

Given the above characteristics, the sample cannot be said to be a probabilistic random sample, but is rather homogeneous for urban, younger respondents, and has greater representation from females than comparable samples obtained from face-to-face interviews in Bangladesh.

The context for the study was the country-wide emergency situation during COVID-19, with required physical distancing recommendations and relevant health regulations in place. Bangladesh was under partial lock-down during the survey period, with only essential services open, offices running on a rotation basis, all educational institutes closed, and public transport services running at 50% capacity. For the purpose of this paper, we use two questions that provide information about trip-making frequency by eight modes of transportation. The modes are **car**, **ride-hailing** (e.g., Uber, Pathao), **rickshaw**, **CNG auto-rickshaw** (a rickshaw-like vehicle powered by compressed natural gas), **bus**, motorcycle/scooter (hereafter just **motorcycle**), **walking**, and **bicycle** (there was an additional catch-all category **other** which we ignore here). Participants in the survey used the following levels to report their frequency of traveling by each mode both before and during COVID-19: _Never_, _Rarely_, _Once a week_, _2-3 trips per week_, _4-5 trips per week_, _Almost daily_.

@@

Figure \ref{fig:column-plot-cases} shows the number of responses (out of 800) in each trip-making frequency class by mode of transportation. The white bars and gray bars are for travel before and during the pandemic, respectively. Considering travel before the pandemic, travel by rickshaw, and bus were relatively common for many respondents (few respondents reported _never_ using these modes). The mode most commonly used on a quotidian basis was walk. In contrast, respondents reported less frequent travel by car, ride-hailing services, CNG auto-rickshaw, motorcycle, and bicycle. During the pandemic, we see that while there were reductions in mobility by car, motorcycle, and bicycle (with more respondents reporting never traveling by these modes), the changes were relatively minor.

```{r column-plot-cases-before-after, echo=FALSE, fig.height= 7, fig.cap="\\label{fig:column-plot-cases}Number of responses by trip-making frequency class and mode, before and during COVID-19"}
freq_by_mode_long %>%
  filter(Mode != "other") %>%
  ggplot(aes(x = frequency, y = Cases, fill = Period)) +
  geom_col(color = "black", position = "dodge") + 
  scale_fill_manual(values = c("before" = "white", "during" = "gray")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  facet_wrap(. ~ Mode, nrow = 4)

# Uncomment to save to jpg files for the publisher
#ggsave(filename = "Figure-1.jpeg", device = "jpeg", height = 9, width = 6.5, units = "in")
```

```{r table-transitions, echo=FALSE}
rbind(car_xtab, 
      rideshare_xtab,
      rickshaw_xtab, 
      cng_xtab,
      bus_xtab, 
      motorcycle_xtab,
      walk_xtab,
      bicycle_xtab) %>% 
  t() %>%
  data.frame() %>% 
  mutate(across(.cols = everything(), function(x) {
    cell_spec(x, "html", bold = TRUE, color = "white", background = spec_color(x, end = 0.8, alpha = 0.3, option = "A", direction = -1))
    })) %>%
  t() %>%
  data.frame() %>%
  rename(Never = X1, Rarely = X2, `Once a week` = X3, `2-3 times per week` = X4, `4-5 times per week` = X5, `Almost daily` = X6) %>%
  rownames_to_column(var = "Before") %>%
  mutate(Before = rep(c("Never", "Rarely", "Once a week", "2-3 times per week", "4-5 times per week", "Almost daily"), 8)) %>%
  kable("html",
        booktabs = TRUE,
        escape = FALSE,
        caption = "\\label{tab:cross-tabulation} Cross-tabulation of counts of travel frequency by mode, before and during COVID-19") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>%
  add_header_above(c(" " = 1, "During" = 6)) %>%
  pack_rows("car", 1, 6) %>%
  pack_rows("ride-hailing", 7, 12) %>%
  pack_rows("rickshaw", 13, 18) %>%
  pack_rows("CNG auto-rickshaw", 19, 24) %>%
  pack_rows("bus", 25, 30) %>%
  pack_rows("motorcycle", 31, 36) %>%
  pack_rows("walk", 37, 42) %>%
  pack_rows("bicycle", 43, 48)  %>%
  footnote(general = "Darker cell colors indicate higher cell values in a row.")
```

The frequency of trip-making by other modes changed more noticeably: the frequency of travel by ride-hailing services, rickshaw, CNG auto-rickshaw, and bus collapsed, with vastly more respondents reporting never using these modes during the pandemic than before. The frequency of walking trips also decreased (fewer respondents report walking almost daily), but the reductions in mobility were not so heavily concentrated at the bottom of the scale.

@@

Despite across-the-board losses of mobility, there appears to have been some adaptation that varied by mode of transportation. To illustrate, `r bus_xtab[6, 1]` respondents, or `r paste(round(bus_xtab[6, 1] / sum(bus_xtab[6,]) * 100, 2), "%", sep="")` of those who traveled by bus almost daily before, reported never using it during the pandemic. In contrast, only `r sum(bus_xtab[1, 2:6])` respondents, or `r paste(round(sum(bus_xtab[1, 2:6]) / sum(bus_xtab[,]) * 100, 2), "%", sep="")` of those who never used buses before started doing so during the pandemic. By way of comparison, `r paste(round(bicycle_xtab[6, 1] / sum(bicycle_xtab[6,]) * 100, 2), "%", sep="")` of respondents who cycled almost daily before the pandemic stopped doing so \- but `r paste(round(sum(bicycle_xtab[1, 2:6]) / sum(bicycle_xtab[,]) * 100, 2), "%", sep="")` who never cycled before started doing so during the pandemic.

@@

# BICYCLE
circle.plot(bicycle_ptab2, "bicycle", nlines = -3)

# Uncomment to save to jpg files for the publisher
#dev.off()
```

The findings reported here suggest important changes in mobility in a developing country during COVID-19, which (given the characteristics of the convenience sample) may be more representative of younger, urban residents in Bangladesh. In general there was a loss of mobility across-the-board, but modes that require interaction with strangers (ride-hailing, bus, rickshaw, and CNG auto-rickshaw) were more affected. Travel by car, motorcycle, and bicycle were somewhat less affected, and there is evidence of some adoption of walking during the pandemic.
